Extended Early Adulthood and Attitude Change

### 2024-06-27 manuscript/formative_period_kk.Rmd
kkedits627

Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991]. In support of this idea, research regularly finds that people in these life stages are more likely to change their preferences, attitudes, and opinions than people in other life stages [@sears2013; @kiley2020]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].

However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

@@

In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially has consequences for the rate at which people undergo cultural changes.

Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of age categories and question. Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) are more likely to make changes of opinion than older respondents in the most recent panels but not in the earliest panels. However, this difference emerged over the second half of 20th Century not because young adults become more likely to make changes, as life-course theories might predict, but because middle-aged and older respondents become less likely to make changes. Our findings complicate the narrative about social change and cultural change and raise questions about what social forces emerged to stabilize attitudes more in middle age for more recent cohorts.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These models of social imprinting have implications for the kinds of patterns we should observe in a population's attitudes over time and suggest specific forms that within-person opinion change should take over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality traits [@wagner2019]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort -- the year of a person's birth - explains much more variance than period -- the year of observation -- for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

Why is it research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

@@

While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader cultural environment that might affect attitude stability. We briefly consider three here.

First, the number of people pursuing and completing higher education increased significantly over the window studied here. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. A population with higher levels of education might be more likely to hold stable opinions regardless of broader social transformations.

Second, we have seen significant expansion of information technology over this window that has dramatically increased access to information about social and political affairs, first through broadcast and cable television and then through the internet and social media. People have greater access to information with which to form opinions.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina; @baldassarri]. People now receive significantly clearer signals from elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020]. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently.

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of an intercept and the duration between observations.

@@

In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations. Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked. In the 1972-76 ANES panel, respondents were asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

Variation in duration between observations within panels lets us separate two kinds of change: change attributable to durable shifts of opinions over time that can be associated with duration between observations and variation that is principally random, or unrelated to the duration between observations. If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

Our theoretical question deals primarily with the amount of change over time ($\beta_0$ and $\beta_1$) changes over time. To explore that, we can add an additional term for time and interact the duration term with time.

$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$

In this formulation, $\beta_2$ captures the change in the baseline level of random change over time. A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time. $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration. A positive $\beta_3$ value would suggest that people are more likely to undergo a durable or persisting change in their opinion over time in later panels than earlier panels.

A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels. As such, we lack large numbers of any particular question, meaning the time trend for any specific question is likely to be unstable. One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels. It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.

Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.

$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3}(d*t) + \epsilon_i $$

@@

In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the item. Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time and duration. These effects are assumed to be normally distributed around a main effect. In practice, we see minimal variation in the question-specific coefficients for the interaction between duration and time, and as such do not allow these to vary at the question level.

For all models, we normalize response scales between 0 and 100.

In all models, respondents in the 18-25 age group serve as the reference category.

@@

@@

Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above. In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for change in recent years, rather than at earlier time points.

The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question. This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$). On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.

@@

Duration between observations is positively but weakly associated with wave-to-wave absolute change. For every additional year between responses, the expected absolute deviance for respondents between the ages of 18 and 25 increases about .9 points. Notably, variance around this main effect is relatively small (s.d. = 0.52), meaning each one year duration only increases the amount changed by between .5 and 1.5. Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023]. While often called "measurement error," this change conflates both random change due to measurement error and true non-persisting change.

The interactions between age and duration are, for the most part, small and not statistically significant, meaning groups make "real" change at roughly the same rates over time. People over 25 make slightly less change than people 18-25 for each additional year of observation, suggesting that the youngest age groups might be making more durable changes of opinion than their older counterparts, but these differences are small and substantively negligible, especially compared to the amount of non-persisting change.

@@

```{r}
new.data %>%
  mutate(group = paste(age_group, name, sep = "-")) %>%
  mutate(year = as.Date("2020-11-12") + dec_diff*3652.5) %>%
  ggplot(aes(x = year, y = yhat, color = age_group)) + 
  geom_hline(yintercept = 0, color = "black") + 
  #geom_line(alpha = .2, aes(group = group)) + 
  geom_smooth(linewidth = 2) + 
  theme_bw() + 
  #facet_wrap(~age_group) + 
  labs(x = "Year", y = "Predicted wave-to-wave change",
       color = "Age Group",
       title = "Predicted wave-to-wave change by age group",
       subtitle = "Individual question trajectories and overall trajectory") + 
  scale_color_brewer(type = "qual", palette = 2) +
  theme(legend.position = "none")
```

@@

Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33). As Americans have pushed back marriage, childbirth, and entering the workforce, we should expect these transitions to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability. We tested that claim by looking at the stability and reliability of political attitude reports among different cohorts as they moved through the same age groups.

The results from the preceding analyses suggest that the changing life course has had minimal impact on rates of attitude change in the population. Despite significant shifts in age of first marriage and age of first child birth, and a much higher proportion of the population attending higher education, attitude behavior as measured by reliability and attitude stability are remarkably similar over time. Across panels, we see consistent evidence that ages 18 to 25 have lower attitude stability and potentially lower attitude reliability relative to the rest of the population, a pattern consistent with the "impressionable years" hypothesis that has broad support in the sociological literature. However, there is no consistent change over time in either reliability or stability estimates for people in this age group. They are different from the rest of the population, but similarly different over time. It does not seem to be the case that the emergence of the "Age of Independence" since the middle of the 20th Century has increased rates of attitude change among people in this group.

>>>>

Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991]. In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].

Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences,” arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

@@

In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course for more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time. Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents’ reported opinions continued to change at roughly the same rate over time.

Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [@]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

Why does research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

@@

While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure. We briefly consider three here.

First, the number of people pursuing and completing higher education increased significantly over the window studied here []. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. A population with higher levels of education might be more likely to hold stable opinions regardless of broader social transformations.

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Both broadcast and cable television and then through the internet and social media. People have greater access to information with which to form opinions.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008]. People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020]. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently. Geographic and associational sorting into lifestyle “enclaves” also reinforce these processes, continually providing clearer signals about which opinions people should hold [].

Broadly speaking, these three forces suggest that opinions . Additionally, peoples’ opinions on issues might

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main interest) or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

To illustrate our approach, consider a one-item, one-panel case (e.g., partisan identification in the 1972-76 panel). We model the absolute change from wave-to-wave as a function of an intercept, the duration between observations in years, age categories, and the interaction between duration and age categories.

@@

In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations. Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked. In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations. If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

We can extend equation to incorporate age groups:

$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(age_{26-33}) + \beta_3(age_{34-64}) + \beta_4(age_{65+}) + \beta_5(d*age_{26-33}) + \beta_6(d*age_{34-64}) + \beta_7(d*age_{65+}) + \epsilon_i $$

In this framework, $\beta_0$ and $beta_1$ now capture the expected change in responses over time for no time between waves and each additional year, respectively, for respondents ages 18-25 (the reference category). $\beta_2$ through $\beta_4$ capture the difference in baseline change between observations for the older groups, and $\beta_5$ through $\beta_7$ capture differences in the effect of an additional year duration for the older age groups.

Our theoretical question deals primarily with whether these short-term rates of change over time themselves change across the broader time span captured across panels. To explore that, we can add an additional term for time and interact all previous coefficients outlined above with time.

$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$

$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to time = 0, which we set to be the most recent observation date (2016). $\beta_8$ estimates the change in the baseline rate of change for the reference group. $\beta_9$ captures change in the effect of duration of observation over time for people in the reference group (ages 18-25). Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.

To this point, we have developed our model for the case of a single question. However, we lack a large number of observations for most items, meaning separate estimates of the time trends for all these items are likely to be inconclusive. One solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels. It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.

Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty. We follow this approach, with the model:

$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$

@@

This model allows for each question its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels

In practice, the model presented in Eq ### is likely too complicated in the face of the data limitations we face.

In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the survey item. Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for duration ($beta_{1j}$), time ($\beta_{2j}$), and their interaction ($\beta_{3j}$). These effects are assumed to be normally distributed around a main effect. In practice, we

@@

## Demographic Variables

To more directly test whether divergences in life course structure drove changes in attitude stability, we include a set of covariates capturing demographic transitions. These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent have ever had a child. We also include an indicator for whether a respondent reported completing college.

To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.

@@

Table XXX presents a comparison of the Bayesian Information Criteria for models allowing for random effects for different coefficients. The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any higher-order interaction. The best-fitting model also omits the third-order interaction between time, age group, and duration of observation. In other words, while there is evidence that age groups changed the amount they changed over time, and that the effect of duration changed over time, there is no evidence that groups experienced differnet rates of change in the rate of durable change.

Model 1 in Table XXX presents the main effects and variance components of the best-fitting multilevel model presented above. In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for the amount of change observed in the most recent years, rather than at earlier time points.

The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question. This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$). On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.

Duration between observations is positively associated with wave-to-wave absolute change. For every additional year between responses, the expected absolute change for respondents increases 1.1 points. As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change. Variance around this main effect is relatively small (s.d. = 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5. Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].

@@

People over 25 make slightly less change than people 18-25 for each additional year of observation, suggesting that the youngest age groups might be making more durable changes of opinion than their older counterparts, but these differences are small and substantively negligible, especially compared to the amount of non-persisting change.

@@

```{r}
new.data %>%
  filter(d == 2) %>%
  mutate(group = paste(age_group, name, sep = "-")) %>%
  mutate(year = as.Date("2020-11-12") + dec_diff*3652.5) %>%
  ggplot(aes(x = year, y = yhat, color = age_group)) + 
  geom_hline(yintercept = 0, color = "black") + 
  geom_line(alpha = .2, aes(group = group)) + 
  geom_smooth(linewidth = 2) + 
  theme_bw() + 
  facet_wrap(~age_group) + 
  labs(x = "Year", y = "Predicted wave-to-wave change",
       color = "Age Group",
       title = "Predicted wave-to-wave change by age group",
       subtitle = "Individual question trajectories and overall trajectory") + 
  scale_color_brewer(type = "qual", palette = 2) +
  theme(legend.position = "none")
```

@@

Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33). Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the 20th Century, we expected these social shifts to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability. We tested that hypothesis by looking at the stability of political and social attitudes among different cohorts as they moved through the same age groups.

The results presented above suggest a more complicated picture than expected. Consistent with expectations, the youngest adults look similar to older adults in the earliest panels. While members of the youngest age categories (18-25 and 26-33) diverged from older members of the population in the rate at which they changed their opinions, showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults. We suggested that this aggregate divergence was driven by counter-vailing trends. First, consistent with our expectations, delays in life course transitions appear to push the youngest members of the population to make more attitude change over time. At the same time, a broad set of social forces worked to stabilize opinions for all members of the population, leading to divergence between the youngest adults and the rest of the population.

The broad pattern of results suggests a broad role for social structures to influence patterns of stable opinion holding. At the aggregate level, broad social currents such as increased levels of education, changes in media environments and access to information, and changes in aggregate social sorting appear to affect …

At another level, social forces that structure the life course appear to produce different rates of change for people in different life stages.
